Reconstructing contact network parameters from viral phylogenies

### 2016-07-03 doc/paper/changes.tex
Some additional paper stuff

Our classification experiments were marginal -- they only estimated one
parameter at a time, when all others were fixed and known. In this situation,
something like k-nearest-neighbours classification or grid search might be just
as effective as ABC, and will certainly be faster. The justification for our
method comes in overcoming the curse of dimensionality when jointly estimating
multiple parameters of a network model. The text added to address the previous
point makes it clearer that the classifier experiments were marginal in nature.
(TODO: We have emphasized this point in the discussion as well: ).

@@

This is a valid point, and in fact drawing $m$ from a distribution is an
approach we considered when we were performing this work. However, since this
is our first attempt to fit network models, we thought it best to stick with
a simple, well-known model rather than defining our own parameters. We did
perform a small experiment allowing heterogeneity among nodes, but in $\alpha$
rather than in $m$, and recovered the average of the two $\alpha$ values.

(TODO: We
have added to the discussion the point that this binary choice is sub-optimal
and suggested this enhancement for future work.)

>>>>

Our classification experiments were marginal -- they only estimated one
parameter at a time, when all others were fixed and known. In this situation,
something like k-nearest-neighbours classification or grid search might be just
as effective as ABC, and will certainly be faster. The justification for our
method comes in overcoming the curse of dimensionality when jointly estimating
multiple parameters of a network model. The text added to address the previous
point makes it clearer that the classifier experiments were marginal in nature.

@@

This is a valid point, and in fact drawing $m$ from a distribution is an
approach we considered when we were performing this work. However, since this
is our first attempt to fit network models, we thought it best to stick with
a simple, well-known model rather than defining our own parameters. We did
perform a small experiment allowing heterogeneity among nodes, but in $\alpha$
rather than in $m$, and recovered the average of the two $\alpha$ values. We
have suggested the possibility of drawing $m$ from a distribution in the
discussion (emphasis is new text): ``More sophisticated models, for example
models incorporating heterogeneity in node behaviour, are likely to provide a
better fit to these data. \emph{For example, rather than using the same $m$ for
all nodes, we could specify a distribution from which each individual
node's $m$ value would be drawn}.''

In addition, we have outlined the difficulties of the binary choice of $m$ in
the discussion of the real data results: ``\ldots the BA model does not allow
any nodes with degree 1 when $m \geq 2$. For this reason, the choice of whether
to allow $m = 1$ in the prior is problematic, as we must choose between an
unrealistic topology (no cycles) and an unrealistic minimum degree.''

### 2016-07-03 doc/paper/paper.Rnw
Some additional paper stuff

Our investigation of published HIV datasets indicated heterogeneity in the
contact network structures underlying several distinct local epidemics. When
interpreting these results, we caution that the BA model is quite simple and
most likely misspecified for these data. In particular, the average degree of a
node in the network is equal to $2m$, and therefore is constrained to be a
multiple of 2. Furthermore, we considered the case $m = 1$, where the network
has no cycles, to be implausible and therefore assigned it zero prior
probability in one set of analyses. This forced the average degree to be at
least four, which may be unrealistically high for sexual networks. The fact
that the estimated values of $\alpha$ differed substantially for three datasets
depending on whether or not $m = 1$ was allowed by the prior is further evidence
of this potential misspecification. However, we note that for \del{two}\add{the
majority} of the datasets, the estimated values of $\alpha$ did not change much
between priors, and the estimates of $I$ were robust to the choice of prior for
all datasets studied (compare \cref{fig:abchpd,fig:abchpdm2}). More
sophisticated models, for example models incorporating heterogeneity in node
behaviour, are likely to provide a better fit to these data.

@@

\add{Unfortunately, the sub-linear region for $\alpha$ identified by both
\textcite{de2007preferential} and \textit{netabc} is also the region of
poorest identifiability (\cref{fig:abcptm2,fig:abcptm3,fig:abcptm4}). This was
reflected in the high level of uncertainty in the estimates, with most 95\%
HPD intervals covering the majority of the range [0, 1]. The value $\alpha =
0.5$ was contained in the 95\% HPD interval for every dataset; consequently, it
is not possible to say with high confidence that any of the $\alpha$ values are
different from each other. In synthetic data, the confidence intervals around
$\alpha$ narrowed when other parameters were marginalized out
(\cref{fig:marginal}). Thus, it is possible that estimates of $\alpha$ could be
made more precise by specifying either exact values or informative priors on
the other BA parameters when these are known.}

\add{I is underestimated in some cases and overestimated in others. Some of the
datasets have most of the posterior mass on m = 1. Estimates of N are
meaningless.}

@@

Our method has a number of caveats, perhaps the most significant being that it
takes a transmission tree as input. In reality, true transmission trees are not
available and must be approximated, often by way of a viral phylogeny. Although
this has been demonstrated to be a fair
approximation~\autocite[e.g.][]{leitner1996accurate}, and is frequently used in
practice~\autocite[e.g.][]{stadler2013uncovering}, the topologies of a viral
phylogeny and transmission tree can differ
significantly~\autocite{ypma2013relating} due to within-host evolution and the
sampling process. In addition, the ABC-SMC algorithm is
computationally intensive, taking about a day when run on 20 cores in parallel
with the settings we described in the methods. Nevertheless, our method is
potentially useful to epidemiological researchers interested in the general
characteristics of the network structure underlying disease outbreaks. This
work, and previous work by our group~\autocite{poon2015phylodynamic}, has
demonstrated that ABC is a broadly applicable and effective framework in
which to perform phylodynamic inference.

>>>>

Our investigation of published HIV datasets indicated heterogeneity in the
contact network structures underlying several distinct local epidemics. When
interpreting these results, we caution that the BA model is quite simple and
most likely misspecified for these data. In particular, the average degree of a
node in the network is equal to $2m$, and therefore is constrained to be a
multiple of 2. Furthermore, we considered the case $m = 1$, where the network
has no cycles, to be implausible and therefore assigned it zero prior
probability in one set of analyses. This forced the average degree to be at
least four, which may be unrealistically high for sexual networks. The fact
that the estimated values of $\alpha$ differed substantially for three datasets
depending on whether or not $m = 1$ was allowed by the prior is further evidence
of this potential misspecification. However, we note that for \del{two}\add{the
majority} of the datasets, the estimated values of $\alpha$ did not change much
between priors, and the estimates of $I$ were robust to the choice of prior for
all datasets studied (compare \cref{fig:abchpd,fig:abchpdm2}). More
sophisticated models, for example models incorporating heterogeneity in node
behaviour, are likely to provide a better fit to these data. \add{For example,
rather than using the same $m$ for all nodes, we could specify a distribution
from which each individual node's $m$ value would be drawn.}

@@

\add{Unfortunately, the sub-linear region for $\alpha$ identified by both
\textcite{de2007preferential} and \textit{netabc} is also the region of
poorest identifiability (\cref{fig:abcpt}). This was reflected in the high
level of uncertainty in the estimates, with most 95\% HPD intervals covering
the majority of the range [0, 1]. The value $\alpha = 0.5$ was contained in the
95\% HPD interval for every dataset; consequently, it is not possible to say
with high confidence that any of the $\alpha$ values are different from each
other. In synthetic data, the confidence intervals around $\alpha$ narrowed
when other parameters were marginalized out (\cref{fig:marginal}). Thus, it is
possible that estimates of $\alpha$ could be made more precise by specifying
either exact values or informative priors on the other BA parameters when these
are known.}

\subsubsection{Other BA parameters}

\add{The true HIV prevalence in a population can be difficult to estimate for
several reasons. HIV-infected individuals may be asymptomatic for months or
years, possibly delaying their awareness of their status. In many contexts, the
risk factors for acquisition of HIV are illegal or stigmatized, which may
represent a barrier to testing, treatment, and/or disclosure of status.
Our simulation study showed that $I$ is weakly identifiable from tree shapes,
however, the estimates of $I$ obtained with \textit{netabc} were upwardly
biased (\cref{fig:abcpt}). In addition, our initial exploratory analysis showed
that the identifiability of $I$ decreases with the number of sampled tips
(\cref{fig:Icrossv}); in real world studies, the proportion of infected
individuals sampled is usually low. The estimated $I$ values for the HIV
datasets ranged from
    X (D)
to
    X (D).}

\add{We were not able to discern and trends toward over- or underestimation of
$I$ from available prevalence data. For example, the authors of the
HET/Botswana data~\autocite{novitsky2013phylogenetic, novitsky2014impact}
estimated that there were 1731 \gls{HIV}-positive individuals in the study
area. \gls{HIV} sequences were obtained from approximately 70\% of these
individuals. The estimated prevalence we obtained was much higher 
    (\Sexpr{d1[dataset == "novitsky2014" & parameter == "I", round(mean)]}),
with a 95\% \gls{HPD} interval spanning nearly the entire prior region. On the
other hand, the study which produced the MSM/USA
data~\autocite{little2014using} enrolled 648 HIV-positive MSM; thus, our
estimate of $I$
    (\Sexpr{d1[dataset == "little2014" & parameter == "I", round(mean)]})
was clearly an underestimate. Post-hoc explanations can be imagined for both of
these results. The HET/Botswana data were collected from a town located
proximally to the country's capital city; the authors suggested that frequent
travel between the two locations may have facilitated linking of their sexual
networks. Thus, the high estimate of \gls{I} we obtained for these data may
include a larger network component based in the capital city. The MSM/USA
result may indicate that the individuals genotyped for the study are members of
a smaller subnetwork which does not include the entire local MSM population.
Unfortunately, none of these hypotheses can be easily tested.}

\add{Over half of the datasets were estimated to have $m = 1$, which produces
tree-like networks without cycles. Since the average degree of a
Barab\'asi-Albert network is $2m$, this value may simply reflect the fact that
most people have a small number of sexual partners, especially when only
recent partnerships are considered~\autocite{liljeros2001web}. In fact, in one
survey, the most common number of partnerships in the past twelve months was
one~\autocite{liljeros2001web}; the BA model does not allow any nodes with
degree 1 when $m \geq 2$. For this reason, the choice of whether to allow $m =
1$ in the prior is problematic, as we must choose between an unrealistic
topology (no cycles) and an unrealistic minimum degree. Extensions to the BA
model which relax this constraint can be imagined and may offer improved
parameter resolution. Estimates of $N$ were not informative for any of the
datasets under either choice of prior, consistent with our simulation results.}

@@

Our method has a number of caveats, perhaps the most significant being that it
takes a transmission tree as input. In reality, true transmission trees are not
available and must be approximated, often by way of a viral phylogeny. Although
this has been demonstrated to be a fair
approximation~\autocite[e.g.][]{leitner1996accurate}, and is frequently used in
practice~\autocite[e.g.][]{stadler2013uncovering}, the topologies of a viral
phylogeny and transmission tree can differ
significantly~\autocite{ypma2013relating} due to within-host evolution and the
sampling process. Progress has very recently been made on this front by
\textcite{giardina2016inference}, who have concurrently developed an ABC-based
method incorporating within-host evolution. The ABC-SMC algorithm is
computationally intensive, taking about a day when run on 20 cores in parallel
with the settings we described in the methods. Nevertheless, our method is
potentially useful to epidemiological researchers interested in the general
characteristics of the network structure underlying disease outbreaks. This
work, and previous work by our group~\autocite{poon2015phylodynamic}, has
demonstrated that ABC is a broadly applicable and effective framework in which
to perform phylodynamic inference.
